---
title: "Toronto Subway Delay Analysis: Causes, Timing, and Line-Specific Patterns"
author: "Avi Walia"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document: 
    fig_width: 10
  github_document:
    df_print: paged
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval=T, echo=F, warning=F, message=F)
```

```{r}
library(opendatatoronto)
library(dplyr)

package <- show_package("996cfe8d-fb35-40ce-b569-698d51fc683b")
resources <- list_package_resources(package)

subway_code_data <- resources |> 
  head(1) |> 
  get_resource()

subway_delay_data <- resources |> 
  tail(1) |> 
  get_resource()

subway_code_data <- subway_code_data |> 
  select("SUB RMENU CODE", "CODE DESCRIPTION...3")
colnames(subway_code_data) <- c("Code", "Code Description")

subway_data <- merge(subway_delay_data, subway_code_data, by = "Code")
```

# Introduction

The Toronto Transit Commission (TTC) subway system is a critical part of Toronto's public transportation network, servicing over a million riders daily across multiple lines. However, subway delays have been a persistent issue, impacting the efficiency and reliability of the service. Understanding the primary causes of these delays and how they vary by time of day and across different subway lines is essential for improving service quality and enhancing passenger experience.

This study aims to investigate the following questions:

1. What are the primary causes of subway delays in Toronto?
2. How do these delays vary by time of day and subway line/station?

**Hypothesis:**

* **Hypothesis 1:** Mechanical issues are the most common causes of subway delays in Toronto.
* **Hypothesis 2:** Delays are more frequent during peak hours (7-9 AM and 4-6 PM) compared to non-peak hours.
* **Hypothesis 3:** The Yonge-University line experiences more frequent delays than other lines due to higher passenger volumes and longer track lengths. Bloor-Yonge and Union stations will have the most and longest delays on this line since they are the most crowded stations in terms of ridership.

To address these questions and test the hypotheses, we will utilize the TTC Subway Delay Data provided by the City of Toronto's Open Data portal. The dataset contains detailed information on delay incidents for the year 2024, including:

* **Delay Codes** and their descriptions indicating the reasons for delays.
* **Time Stamp Information** such as date, time, and day of the week.
* **Location Details** including the station and subway line affected.
* **Duration of Delays** measured in minutes.

The data was acquired using the City's Open Data API. By exploring and analyzing this dataset, we seek to identify patterns and trends in delay causes, assess peak times for delays, and determine if specific lines and/or stations are more prone to certain types of delays. This analysis will inform potential strategies for mitigating delays and improving the TTC's operational efficiency.




